some title

### 2018-01-08 2018/paper.tex
partial correlation

\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator{\sVar}{sVar}
\DeclareMathOperator{\card}{card}
\DeclareMathOperator{\Cov}{Cov}
\DeclareMathOperator{\Corr}{Corr}
\DeclareMathOperator{\sCov}{sCov}
\DeclareMathOperator{\sCorr}{sCorr}
\DeclareMathOperator{\E}{E}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\I}{\mathbb{I}} % индикатор события
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{lemma}[theorem]{Lemma}

@@

\begin{figure}[h!]
\begin{center}
\includegraphics{images/corr_def.pdf}
\caption{Vector $x$ of length $\sigma_x$ and random vector $y$ of length $\sigma_y$, $\cos \varphi$ is the angle between $x$ and $y$.}
\label{fig:corr_def}
\end{center}
\end{figure}

@@

\begin{figure}[h!]
\begin{center}
\includegraphics{images/corr_proj.pdf}
\caption{Vector $y$ projected onto vector $x$.}
\label{fig:corr_proj}
\end{center}
\end{figure}

@@

\begin{proof}
Firstly, we project vectors $x$ and $y$ onto $Lin^{\perp}(\mathbf{1})$ in order to get $x^c = x - \bar x$ and $y^c = y - \bar y$
(`c' stands for `centred'). It can be shown that the matrix corresponding to projecting onto the line spanned by a vector of all ones has the following form
\[
\frac{\mathbf{1}^T \mathbf{1}}{\mathbf{1} \mathbf{1}^T} = \frac{\begin{pmatrix} 1 & \ldots & 1 \end{pmatrix} \begin{pmatrix} 1 \\ \ldots \\ 1 \end{pmatrix}}{\sum_{i=1}^n 1} = \begin{pmatrix} \frac{1}{n} & \ldots & \frac{1}{n} \\ \vdots & \ddots & \vdots \\ \frac{1}{n} & \ldots & \frac{1}{n} \end{pmatrix}
\]
Thus, projecting onto an orthogonal subspace is equivalent to substracting the projected vector, i.e., the vector of averages, from the original one.

@@

\begin{figure}
  \centering
  \includegraphics{images/corr_const_cent.pdf}
  \caption{Centred vectors $x^c$ and $y^c$}
  \label{fig:corr_xyc}
\end{figure}

@@

\begin{figure}[h!]
\center{\includegraphics{images/corr_eq_positive.pdf}}
\caption{Vectors $x$, $y$ and $\mathbf{1}$.}
\label{fig:corr_pos_base}
\end{figure}

@@

\begin{figure}[h!]
\begin{center}
\subfigure[]{
\includegraphics[width=0.4\linewidth]{images/corr_eq_positive_xy_centered.pdf}
\label{fig:corr_pos_xyc}}
\hspace{4ex}
\subfigure[]{
\includegraphics[width=0.4\linewidth]{images/corr_eq_positive_yhat.pdf}
\label{fig:corr_pos_yhatc}}
\caption{\subref{fig:corr_pos_xyc}: `Centred' $x$ and $y$, i.e., projected onto $Lin^{\perp}(\mathbf{1})$; \subref{fig:corr_pos_yhatc}: `Centred' $\hat y$, i.e., projected onto $Lin^{\perp}(\mathbf{1})$.}
\end{center}
\end{figure}

@@

\begin{theorem}
The point of averages lies on the estimated regression line.
\end{theorem}

@@

%\section{Partial correlation}

>>>>

\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator{\sVar}{sVar}
\DeclareMathOperator{\pVar}{pVar}
\DeclareMathOperator{\card}{card}
\DeclareMathOperator{\Cov}{Cov}
\DeclareMathOperator{\Corr}{Corr}
\DeclareMathOperator{\sCov}{sCov}
\DeclareMathOperator{\pCov}{pCov}
\DeclareMathOperator{\sCorr}{sCorr}
\DeclareMathOperator{\pCorr}{pCorr}
\DeclareMathOperator{\E}{E}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\I}{\mathbb{I}} % индикатор события
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{lemma}[theorem]{Lemma}
\theoremstyle{definition}
\newtheorem{definition}{Definition}

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[width=0.4\linewidth]{images/corr_def.pdf}
\caption{Vector $x$ of length $\sigma_x$ and random vector $y$ of length $\sigma_y$, $\cos \varphi$ is the angle between $x$ and $y$.}
\label{fig:corr_def}
\end{center}
\end{figure}

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[width=0.4\linewidth]{images/corr_proj.pdf}
\caption{Vector $y$ projected onto vector $x$.}
\label{fig:corr_proj}
\end{center}
\end{figure}

@@

\begin{proof}
Firstly, we project vectors $x$ and $y$ onto $Lin^{\perp}(\mathbf{1})$ in order to get $x^c = x - \bar x$ and $y^c = y - \bar y$
(`c' stands for `centred'). It can be shown that the matrix corresponding to projecting onto the line spanned by a vector of all ones has the following form
\[
\frac{\mathbf{1}^T \mathbf{1}}{\mathbf{1} \mathbf{1}^T} = \frac{\begin{pmatrix} 1 & \ldots & 1 \end{pmatrix} \begin{pmatrix} 1 \\ \vdots \\ 1 \end{pmatrix}}{\sum_{i=1}^n 1} = \begin{pmatrix} \frac{1}{n} & \ldots & \frac{1}{n} \\ \vdots & \ddots & \vdots \\ \frac{1}{n} & \ldots & \frac{1}{n} \end{pmatrix}
\]
Thus, projecting onto an orthogonal subspace is equivalent to substracting the projected vector, i.e., the vector of averages, from the original one.

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.45\linewidth]{images/corr_const_cent.pdf}
  \caption{Centred vectors $x^c$ and $y^c$}
  \label{fig:corr_xyc}
\end{figure}

@@

\begin{figure}[h!]
\center{\includegraphics[width=0.35\linewidth]{images/corr_eq_positive.pdf}}
\caption{Vectors $x$, $y$ and $\mathbf{1}$.}
\label{fig:corr_pos_base}
\end{figure}

@@

\begin{figure}[h!]
\begin{center}
\subfigure[]{
\includegraphics[width=0.6\linewidth]{images/corr_eq_positive_xy_centered.pdf}
\label{fig:corr_pos_xyc}}
%\hspace{4ex}
\subfigure[]{
\includegraphics[width=0.6\linewidth]{images/corr_eq_positive_yhat.pdf}
\label{fig:corr_pos_yhatc}}
\caption{\subref{fig:corr_pos_xyc}: `Centred' $x$ and $y$, i.e., projected onto $Lin^{\perp}(\mathbf{1})$; \subref{fig:corr_pos_yhatc}: `Centred' $\hat y$, i.e., projected onto $Lin^{\perp}(\mathbf{1})$.}
\end{center}
\end{figure}

@@

\begin{theorem}
In a linear regression model with one explanatory variable and constant term
\[
y = \beta_1 + \beta_2 x + \varepsilon
\]
the point of averages lies on the estimated regression line.
\end{theorem}

@@

\section{Partial correlation}

\subsection{Definition of partial correlation}

\marginnote{
Partial correlation is the measure of degree of dependence between two random
variables while controlling for the effect of other random variables:
\[
\pCorr(X,Y; Z) = \frac{\pCov(X,Y;Z)}{\sqrt{\pVar(X;Z)\pVar(Y;Z)}}
\]
where $\pVar(X;Z) = \Var(X - \alpha Z)$, $\alpha$ is such a constant that $\Cov(X - \alpha Z, Z) = 0$,
and $\pCov(X,Y; Z) = \Cov(X - \alpha Z, Y - \beta Z)$, $\alpha$, $\beta$ are such
constants that $\Cov(X - \alpha Z, Z) = \Cov(Y - \beta Z, Z) = 0$
}

Partial correlation can be defined into two ways.
We will provide both definitions and show their equivalence.

\begin{definition}
Partial correlation between random variables $X$ and $Y$ holding random variable $Z$
fixed is the correlation coefficient between the residuals in regression of $X$ onto
$Z$ and the residuals in regression of $Y$ onto $Z$.
\end{definition}

Firstly, we project random variable $X$ onto $Z$, which yields $\E(X)$.
The residuals in this regression are $X - \E(X)$ — a vector in $Lin^{\perp}(Z)$.
We will call this variable `centered' and label it as $\widetilde X$.
Repeating this step for $Y$ yields `centerd' variable $\tilde Y = Y - \E(Y) \in Lin^{\perp}(Z)$.
The angle between $\widetilde X$ and $\widetilde Y$ ($\varphi$ in Figure~\ref{fig:pcorr_def1})
is the correaletion coefficient between these `centred' random variables and
the partial correlaiton between the original ones.

\begin{figure}
  \centering
  \includegraphics[width=0.55\linewidth]{images/partial_corr.pdf}
  \caption{Partial correlation between $X$ and $Y$ while $Z$ is fixed.}
  \label{fig:pcorr_def1}
\end{figure}

\begin{definition}
Partial correlaiton between random variables $X$ and $Y$ holding random variable $Z$
fixed is the geometric mean between the coefficient $\hat \beta_{XY}$ in regression
\[
\hat X = \hat \beta_{XY} Y + \hat \beta_{XZ} Z
\]
and the coefficinent $\hat \beta_{YX}$ in regression
\[
\hat Y = \hat \beta_{YX} X + \hat \beta_{YZ} Z
\]
which has the same sign as the coefficients $\hat \beta_{XY}$ and $\hat \beta_{YX}$.
\end{definition}

Following the definition, we need to start with regressing variable $X$ onto
$Y$ and $Z$. Then, the vector we obtained $\hat X$ can be broken up into the sum of
$\hat \beta_{XY} Y$ and $\hat \beta_{XZ} Z$.
Projecting $\hat \beta_{XY} Y$ onto $Lin^{\perp}(Z)$ we results in a vector
$\alpha \widetilde Y$ where $\widetilde Y = Y - \E(Y)$ is the projection of $Y$ onto $Lin^{\perp}(Z)$.

By the properties of similar triangles
\[
\frac{\beta_{XY} Y}{Y} = \frac{\alpha \widetilde Y}{\widetilde Y} \Leftrightarrow \beta_{XY} = \alpha
\]

In the same way we perform a regression of $Y$ onto $X$ and $Z$ and repeat the
same steps as for $X$. Finally, we get the whole picture:

\begin{figure}[ht!]
\begin{center}
\subfigure[]{
\includegraphics[width=0.55\linewidth]{images/part_corr_2.pdf}
\label{fig:part_corr_alt}}
%\hspace{4ex}
\subfigure[]{
\includegraphics[width=0.35\linewidth]{images/part_corr_2_lin.pdf}
\label{fig:part_corr_alt_lin}}
\caption{\subref{fig:part_corr_alt}: Alternative definition of the partial correlation;
\subref{fig:part_corr_alt_lin}: $Lin^{\perp}(Z)$.}
\end{center}
\end{figure}

Having plotted $Lin^{\perp}(Z)$ now we can express $\cos \varphi$ in terms of $\beta_{XY}$
and $\beta_{YX}$

\begin{equation}\label{eq:part_cor_cos}
\begin{split}
\cos \varphi &= \frac{\vert \beta_{XY} \widetilde Y \vert}{\vert \widetilde X \vert} = \vert \beta_{XY} \vert \\
\cos \varphi &= \frac{\vert \beta_{YX} \widetilde X \vert}{\vert \widetilde Y \vert} = \vert \beta_{YX} \vert \\
\cos^2 \varphi &= \vert \beta_{XY} \beta_{YX} \vert \stackrel{sign(\beta_{XY}) = sign(\beta_{YX})}{=} \beta_{XY} \beta_{YX}
\end{split}
\end{equation}

Recall that the angle $\varphi$ can be interpreted as the partial correlation
between $X$ and $Y$ holding $Z$ fixed, so it follows form equaitions~\eqref{eq:part_cor_cos}

\[
\pCorr^2(X,Y; Z) = \cos^2 \varphi = \beta_{XY} \beta_{YX}
\]
